---
title: "ggplot Piper Diagram"
date: "January 24, 2018"
output: github_document
---
**This is a fork of the gist 
https://gist.github.com/johnDorian/5561272 (by Jason Lessels, jlessels@gmail.com).**
Forking a gist into a git-repository unfortunately does not preserve the relation to the forked project, sorry for that.

Jasons comments:

>A piper diagram based on the ternary plot example here: http://srmulcahy.github.io/2012/12/04/ternary-plots-r.html.
*(this link is broken, Note from Marko, Jan 2018)*

>This was written quickly, and most likely contains bugs - I advise you to check it first. This now consists of two functions. *transform_piper_data()* transforms the data to match the coordinates of the piper diagram. *ggplot_piper()* does all of the background.


```{r}
source("ggplot_Piper.R")
library("hydrogeo")
```

# Example
## Data input

Input data need to be percent of meq/L !  
*meq/L = mmol/L \* valence* ([Wikipedia: Equivalent (chemistry)](https://en.wikipedia.org/wiki/Equivalent_(chemistry))) with

element | valence
--------|--------
Ca      | 2
Mg      | 2
Na      | 1
K       | 1
Cl      | 1
SO4     | 2
CO3     | 2
HCO3    | 1

### Example data

```{r data}
milliequivalents <- list(Ca   = c(43, 10, 73, 26, 32),
                         Mg   = c(30, 50, 83, 14, 62),
                         Na   = c(54, 76, 3, 14, 12),
                         K    = c(31, 22, 32, 22, 11),
                         Cl   = c(24, 10, 12, 30, 43),
                         SO4  = c(24, 10, 12, 30, 43),
                         CO3  = c(24, 10, 12, 30, 43),
                         HCO3 = c(42, 110, 12, 3, 4),
                         "WaterType" = c(2, 2, 1, 2, 3),
                         "SecondFactor" = c("low", "low", "high", "high", "high"),
                         IDs = c("A","B","C","D","E") )
percents <- toPercent(milliequivalents)

data <- as.data.frame(percents)

data
```


### Check...
 ... should add up to 100%

```{r check}
cation.sums <- apply(data[, c("Ca", "Mg", "Na", "K")], 1, FUN = sum)
anion.sums  <- apply(data[, c("Cl", "SO4", "CO3", "HCO3")], 1, FUN = sum)

cation.sums
anion.sums
```

## Transformation

```{r transformed}
piper_data <- transform_piper_data(Ca   = data$Ca,
                                   Mg   = data$Mg,
                                   Cl   = data$Cl,
                                   SO4  = data$SO4,
                                   name = data$IDs)
 piper_data <- merge(piper_data,
                     data[, c("WaterType", "SecondFactor", "IDs")],
                     by.y = "IDs",
                     by.x = "observation")

piper_data
```

## Plot

The piper function now just plots the background

```{r basePlot, fig.width=10, fig.asp=1}
ggplot_piper()
```

Now points can be added like...

```{r withPoints, fig.width=10, fig.asp=1}
ggplot_piper() + geom_point(aes(x,y), data=piper_data)
```

... colouring the points can be done using the observation value

```{r withColouredPoints, fig.width=10, fig.asp=1}
ggplot_piper() + geom_point(aes(x,y, colour=factor(observation)), data=piper_data)
```

The size can be changed like..

```{r withColouredResizedPoints, fig.width=10, fig.asp=1}
ggplot_piper() + geom_point(aes(x,y, colour=factor(observation)), size=4, data=piper_data)
```

Grouping by multiple factors:

```{r multiple groups, fig.width=10, fig.asp=1}
ggplot_piper() + geom_point(aes(x,y,
                                color = factor(WaterType),
                                shape = SecondFactor),
                            size = 4, stroke = 2, data = piper_data) +
  scale_shape_manual(values = c(21:26)) +
  theme(legend.position = "top", legend.text = element_text(color = "red", size = 20))
```

Advanced example:

```{r advanced, fig.width=10, fig.asp=1}
ggplot_piper() + 
  geom_point(aes(x, y,
                 colour = factor(WaterType),
                 shape  = factor(WaterType)), 
             size=4, data = piper_data) + 
  scale_colour_manual(name="legend name must be the same",
                      values=c("#999999", "#E69F00", "#56B4E9"),
                      labels=c("Control", "Treatment 1", "Treatment 2")) +
  scale_shape_manual(name="legend name must be the same", values=c(1, 2, 3, 4 ,5), labels=c("Control", "Treatment 1", "Treatment 2")) +
  theme(legend.position = c(.8, .9))
```




